---
title: "bootstrap"
author: "Daniel Hill"
format: pdf
editor: visual
---

## Bootstrap sampline

in this document we are going to perform bootstrap sampling on the clean data set and create a new csv file containing the bootstrapped data that we can use in further tasks. I would rather create on file with bootstrap data instead of looping through the vectors and calculating the statistics for each one every time like in the simulation.qmd. I think the result should be the same, but code should be cleaner to write this way.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  fig.width = 6,      
  fig.height = 4,     
  out.width = "80%",  
  fig.align = "center"
)
```


```{r}
library(tidyverse)
set.seed(456)
```

```{r}

data <- read_csv("data_clean.csv")

bootstrap_data <- data %>% sample_n(size=10000, replace=TRUE)

original_dist <- data %>% count(label)
bootstrap_dist <- bootstrap_data %>% count(label)


original_dist
bootstrap_dist
```

```{r}
original_dist <- original_dist %>% mutate(percentage = n / nrow(data) )
bootstrap_dist <- bootstrap_dist %>% mutate(percentage = n / nrow(bootstrap_data) )

original_dist
bootstrap_dist 
```
looks good to me


```{r}
write_csv(bootstrap_data, "bootstrap_data.csv")
```


